Demographic Species Distribution Models

### 2021-05-12 manuscript/appendix_a.Rmd

@@

An approach similar to this can also used be used to compute relative abundances from a static transition matrix for a  matrix metapopulation model - in that model the stable distribution between sites already provides a relative abundance estimate, so there is no need for a model to relate the intrinsic growth rate to relative abundance. See [*Dispersal with a static transition matrix*](#static_dispersal) for details.

@@

In static transition models considered above, carrying capacity is modelled as some function of the intrinsic growth rate, an relationship which implicitly represents density dependence. Rather than considering that density dependence isa  function of the growth rate, it may be more ecologically realistic to explicitly model the vital rates as being density-dependent - that is having the survival probabilities or fecundity rates at a given time point depend on the population density at the previous time point. This enables incorporation of more detailed ecological knowledge about how vital rates drive abundance, which may be important to consider if the answer to the ecological question of interest depends on density dependence.

Density dependent vital rates can be incorporated in a DSDM by constructing the dynamic population transition matrix $\mathbf{A}_{i,t}$, from dynamic vital rates that are computed as a function of both the environment and the total population (i.e. across all life stages $M$) in the previous time point. For example, these density-dependent relationships could be a linear function of population density - controlled by parameters $\alpha_F$ and $\alpha_S$ - on the link-scale of the vital rates. Where $\alpha_F$ and $\alpha_S$ are negative, density-dependence will lead to self-regulating populations that will ultimately converge on a finite carrying capacity.

@@

\begin{align}
\Lambda_i &=  \mathbf{n}_{i, T} \\ \nonumber
\mathbf{n}_{i,t+1} &= \mathbf{A}_{i,t} \mathbf{n}_{i,t} \\ \nonumber
\mathbf{A}_{i,t} &= \begin{bmatrix} F_{i,t} \cdot S_{J_{i,t}} & F_{i,t} \cdot S_{A_{i,t}} \\ S_{J_{i,t}} & S_{A_{i,t}}  \end{bmatrix}  \\ \nonumber
ln(F_{i,t}) &= \mathbf{X}_i \beta_F + \alpha_F \textstyle\sum_{j=1}^M{\mathbf{n}_{i,j,t}}\\ \nonumber
logit(S_{A_{i,t}}) &= \mathbf{X}_i \beta_S + \alpha_S \textstyle\sum_{j=1}^M{\mathbf{n}_{i,j,t}} \\ \nonumber
logit(S_{J_{i,t}}) &= logit(S_{A_{i,t}}) - \gamma
\end{align}
where $\alpha_F$ and $\alpha_S$ are negative scalar parameters controlling the strength of density dependenceadnt he detection rate, and $T$ is positive integer giving a number of iterations chosen such that the population sizes have converged to withing a tolerance, i.e. $|\mathbf{n}_{i,T} - \mathbf{n}_{i,T-1}| / \mathbf{n}_{i,T-1} < \epsilon$ for some small positive tolerance $\epsilon$. This criterion means that the value of the initial population sizes $\mathbf{n}_{i,0}$ do not impact upon the model.

@@

Where the data or questions of the analysis are such that equilibrium population sizes are not representative of the population, the density-dependent vital rate model above can be iterate for predetermined number of timepoints $T$, such that each time $t$ corresponds to an observed time in the distribution data. For example, where the impacts of short-term variation in environmental conditions (environmental stochasticity) on population sizes is important, the transition matrix $\mathbf{A}_{i,t}$ at each time can be computed as a function of the design matrix of observed environmental conditions $\mathbf{X}_t$ at that time, and the current population density. The resulting population size $\mathbf{n}_{i,t}$ at each time $t$ can then be used in an observation model for the observed distribution data at that time $y_{i,t}$.

>>>>

**BP: Here and Ricker: should A be an NxK matrix, as in first section, above?; Further below, M seems to now be used for number of stages rather than densitiy dependence parameter**

@@

An approach similar to this can also used be used to compute relative abundances from a static transition matrix for a  matrix metapopulation model: in that model the stable distribution between sites already provides a relative abundance estimate, so there is no need for a model to relate the intrinsic growth rate to relative abundance. See [*Dispersal with a static transition matrix*](#static_dispersal) for details.

@@

In static transition models considered above, carrying capacity is modelled as some function of the intrinsic growth rate; a relationship which implicitly represents density dependence. Rather than considering that density dependence is a function of the growth rate, it may be more ecologically realistic to explicitly model the vital rates as being density-dependent: that is, having the survival probabilities or fecundity rates at a given time point depend on the population density at the previous time point. This enables incorporation of more detailed ecological knowledge about how vital rates drive abundance, which may be important to consider if the answer to the ecological question of interest depends on density dependence.

Density dependent vital rates can be incorporated in a DSDM by constructing the dynamic population transition matrix $\mathbf{A}_{i,t}$, from dynamic vital rates that are computed as a function of both the environment and the total population (i.e. across all life stages $M$) in the previous time point. For example, these density-dependent relationships could be a linear function of population density, controlled by parameters $\alpha_F$ and $\alpha_S$ on the link-scale of the vital rates. Where $\alpha_F$ and $\alpha_S$ are negative, density-dependence will lead to self-regulating populations that will ultimately converge on a finite carrying capacity.

@@

\begin{align}
\Lambda_i &=  \mathbf{n}_{i, T} \\ \nonumber
\mathbf{n}_{i,t+1} &= \mathbf{A}_{i,t} \mathbf{n}_{i,t} \\ \nonumber
\mathbf{A}_{i,t} &= \begin{bmatrix} F_{i,t} \cdot S_{J_{i,t}} & F_{i,t} \cdot S_{A_{i,t}} \\ S_{J_{i,t}} & S_{A_{i,t}}  \end{bmatrix}  \\ \nonumber
ln(F_{i,t}) &= \mathbf{X}_i \beta_F + \alpha_F \textstyle\sum_{j=1}^M{\mathbf{n}_{i,j,t}}\\ \nonumber
logit(S_{A_{i,t}}) &= \mathbf{X}_i \beta_S + \alpha_S \textstyle\sum_{j=1}^M{\mathbf{n}_{i,j,t}} \\ \nonumber
logit(S_{J_{i,t}}) &= logit(S_{A_{i,t}}) - \gamma
\end{align}
where $\alpha_F$ and $\alpha_S$ are negative scalar parameters controlling the strength of density dependence and the detection rate, and $T$ is positive integer giving a number of iterations chosen such that the population sizes have converged to within a tolerance, i.e. $|\mathbf{n}_{i,T} - \mathbf{n}_{i,T-1}| / \mathbf{n}_{i,T-1} < \epsilon$ for some small positive tolerance $\epsilon$. This criterion means that the value of the initial population sizes $\mathbf{n}_{i,0}$ do not impact upon the model.

@@

Where the data or questions of the analysis are such that equilibrium population sizes are not representative of the population, the density-dependent vital rate model above can be iterated for predetermined number of timepoints $T$, such that each time $t$ corresponds to an observed time in the distribution data. For example, where the impacts of short-term variation in environmental conditions (environmental stochasticity) on population sizes is important, the transition matrix $\mathbf{A}_{i,t}$ at each time can be computed as a function of the design matrix of observed environmental conditions $\mathbf{X}_t$ at that time, and the current population density. The resulting population size $\mathbf{n}_{i,t}$ at each time $t$ can then be used in an observation model for the observed distribution data at that time $y_{i,t}$.

### 2021-05-12 manuscript/ms.Rmd

3. Example applications mapping the Carolina Chickadee in North Ameerica, and a Protea in southern Africa indicate that DSDMs have similar predictive performance to SDMs, and produce plausible spatial estimates of vital rates. Even when DSDMs are provided little information on the ecology of the target species, predictions of distributions are well-identified, and increased uncertainty about population dynamics is reflected in more uncertain estimates of vital rates.

@@

Whilst users typically have some knowledge of the processes driving these aspects of species distributions - such as the species' life history, approximate survival and fecundity rates or dispersal ability - incorporating this knowledge into correlative statistical models is problematic.
On the other hand, ecologists rarely have enough information to accurately predict species distributions from their knowledge of the species life history and demographics alone, since quantifying how vital rates vary over space typically requires large-scale long-term field studies [@merowUsingIntegralProjection2014] or detailed pyhsiological modelling [@kearneyModellingSpeciesDistributions2008].

@@

A number of models have been made to improve the realism in species distribution models, whilst still enabling them to be fitted to the distribution data that are commonly available.
For example, dynamic range models (DRMs) extend correlative SDMs by explicitly modelling dispersal processes and spatial variation in population growth rates [@pagelForecastingSpeciesRanges2012].
DRMs model population growth rates using a correlative model and environmental covariates, just as a correlative SDM models the abundance or probability of presence of a species. 
DRMs can also be fitted to distribution data, though they are intended for use with timeseries of abundance or presence-only data, since repeat observations are needed to accurately quantify population growth rates and dispersal.

Because they do not model vital rates directly DRMs, like correlative SDMs, are unable to predict the impact on distributions of a change - for example a conservation intervention like predator control - that affects a single vital rate.

@@

By contrast, matrix population models and integral projection models (IPMs) are demographic models of single populations that construct and analyse a transition matrix representing key demographic processes [@merowAdvancingPopulationEcology2014].
Spatial IPMs extend this idea to map distributions by modelling demographic rates as a function of covariates, then analysing transition matrices under different conditions to map emergent demographic quantities (such as the intrinsic growth rate) to characterize the species’ distribution [@merowUsingIntegralProjection2014].

@@

where $\mathbf{X}$ is a design matrix of environmental covariates and an intercept dummy variable for the fecundity and survival submodels and $\beta_F$ and $\beta_S$ are vectors of regression coefficients for the two submodels.
Rather than separately modelling spatial variation in adult and juvenile fecundity, I assume that the environmental predictors of survival are the same in both stages, but that juvenile survival differs from adult survival by a constant amount on the logit scale - to maintain the (0,1) constraint on survival probabilities.

@@

where $eigval()$ denotes the dominant eigenvalue of a symmetric matrix, yielding the intrinsic growth rate $\lambda_i$ of the populations at location $i$. The theoretical carrying capacity at each site $K_i$ is computed from the carrying capacity relationship of the Beverton-Holt model. $K_i$ has negative values $\lambda_i < 1$, so it is transformed to an always-positive relative abundance parameter $\Lambda_i$ via the softplus function. The softplus transformation is a smooth version of the Heaviside function (which sets negative values to zero), so $\Lambda_i$ closely approximates $K_i$ for most $lambda_i > 1$, and rapidly approaches zero below $lambda_i = 1$. This formulation is statistically convenient, but as discussed in the text below, DSDMs can incorporate more complex and realistic forms of density-dependence if required. 
The parameter $\alpha$ controls the relationship between the intrinsic population growth rate and relative abundance. It is a compound of two parameters that are only identified together: the Beverton-Holt density-dependence parameter relating growth rates to carrying capacities; and the detection rate - the fraction of individuals in the total population that are detected during a presence-absence survey at site $i$.  
From $\Lambda_i$, we obtain the probability $p_i$ of detecting (one or more individuals of) the species $y_i$, under a poisson sampling distribution [@pagelQuantifyingRangewideVariation2014].

@@

The spatially-averaged fecundity and adult survival estimates from @ryuDevelopingPopulationModels2016 were used to define priors on regression coefficients for the vital rate submodels. These estimated vital rates correspond to populations at the $J$ survey stations used in the recapture analysis (station names provided in the supplemental data to @ryuDevelopingPopulationModels2016) - locations where the species occurs. I defined a prior on $\beta_F$ and $\beta_S$ such that values were *a priori* more likely if they resulted in estimates of $F_j$ and $S_{A_j}$, for each survey station location $j$, that were more consistent with the estimates in @ryuDevelopingPopulationModels2016.
First, the means and variances of these estimates were used to calculate equivalent parameters of univariate normal distributions on the link-scale of the regression submodels, resulting in the implied priors: $ln(F_j) \sim N(-0.63, 0.18)$, $logit(S_{A_j}) \sim N(1.12, 1.28)$, and $logit(S_{J_j}) \sim N(-0.05, 0.85)$.
These normally-distributed implied priors correspond exactly to multivariate normal priors on the vectors of regression coefficients, since the matrix-multiplication that maps between regression coefficients and link-scale response is an affine transformation, under which the multivariate normal distribution is closed. The parameters of the priors over $\beta_F$ are therefore given by: $\mu_F = \mathbf{X}_*^{-1}(\text{-}0.63\, \mathbf{I})$ and $\mathbf{\Sigma}_F = \mathbf{X}_*^{-1} (0.18\,\mathbf{I}) \mathbf{X}_*^{-T}$, where $\mathbf{X}_*$ denotes the design matrix over the $J$ survey stations, $\mathbf{X}_*^{-1}$ its Moore-Penrose pseudoinverse, $\mathbf{X}_*^{-T}$ the transpose of that pseudoinverse, and $\mathbf{I}$ an identity matrix of dimension $J$. The prior parameters $\mu_S$ and $\mathbf{\Sigma}_S$ were defined in the same way using the implied prior parameters over $logit(S_{A_j})$: 1.12, and 1.28.
Note that these priors do not provide any information on the direction of the relationships between vital rates and environmental covariates (the intercept elements of $\mu_F$ and $\mu_S$ are informed, but the prior mean for the other terms is $\approx 0$), but by providing information on plausible magnitudes of, and correlations between these parameters the prior favours regression equations that pass close to the vital rate estimates at $F_j$ and $S_j$.
The prior for $\gamma$, the logit-difference between adult and juvenile survival, is given by the difference between the implied normal priors on logit-survival of adults and juveniles, which is also normal.
In the absence of any meaningful prior information for the competition effects or detection rates, I assigned $\alpha$ an improper uniform prior on all positive values.

I fitted the model to a random subset of `r bbs_n_train` presence-absence records from the North American Breeding Bird Survey in 2015 [@sauerNorthAmericanBreeding2011]. I used five environmental covariates in $\mathbf{X}$: percentage cover of deciduous forest, coniferous forest, and mixed forest (to represent breeding habitat for this species) mean temperature (to capture thermal limits on survival of the species), and annual precipitation (as a proxy for insect abundance and plant growth, food sources for this species). The forest cover layers were obtained from the maxlike R package [@royleLikelihoodAnalysisSpecies2012], and the climatic covariates were bioclim layers 1 and 12 from the WorldClim 2.0 dataset [@fickWorldClimNew1km2017]. In general, ecological knowledge could be used to select separate covariates to use for each of the submodels. However, in the absence of a strong belief that any of these covariates would affect only one vital rate, all six covariates were used in both submodels. These covariate values were centred and scaled (before calculation of the priors on $\beta_F$ and $\beta_S$) to enable comparison of regreesion coefficients as effect sizes.

@@

Note that this model could undoubtedly be improved by including more information of the species ecology and vital rates - for example constraining spatial estimates of fecundity and survival to within a range of plausible values - via prior distributions or changes to model structure. As with any model, the reliability of these vital rate and distribution predictions will depend on the assumptions made and information provided by the modeller. Unlike in purely correlative models however, these assumptions are made clear by explicitly considering spatially-varying vital rates.

@@

Whilst statistical inference against distribution data could be performed on almost any demographic model, I restrict the models discussed here to those that are amenable to efficient MCMC methods. Specifically: models with likelihoods that are tractable (the exact value of likelihood can be computed from the parameters and data), and can be differentiated with respect to parameter values. This enables the use of modern gradient-based MCMC samplers [@monnahanFasterEstimationBayesian2017], thought it necessarily restricts the scope of these models to those with only continuous parameters. The model variations I propose below encompass a wide range of features whist remaining within these constraints, and can also make use of efficient vectorised and linear algebra routines, meaning that should scale well to the large datasets commonly used for fitting SDMs.

@@

Density dependence can be represented in DSDMs by modelling demographic rates as a function of the current population state [@ellnerIntegralProjectionModels] (the abundances of each life stage) as well as covariates.
For example, the current density of individuals could determine fecundity for the next iteration of the dynamics. The matrix iterations will then converge to a stable state; the population carrying capacity or the expected abundance. this cna be achieved by iterating populations a sufficient number of time $K$ for the populations to converge on stable population sizes.

@@

for iteration $k$ in $K$.
 note that this can be combined with Eq 5, or can be iterated for a sufficient number of timesteps (T) to get a temporally-static density-dependent model. In practice, T can be determined adaptively, iterating a sufficient number of times for N_i to converge and then terminating.

@@

For example, theese linear models could be applied to very large numbers of covariates and transformations thereof in combination with regularising priors to perform automatic variable selection, similarly to the popular Maxent SDM software[@phillipsModelingSpeciesDistributions2008]. Or to specify mixed-effects models to account for clustered sampling designs, or use random slopes to share environmental reponse information between species in a biotic interaction DSDM. Instead of linear models, non-linear functions could be introduced with generalisd additive models or Gaussian process models, for which there are existing greta extension packages: greta.gam and greta.gp [@goldingGretaGamGeneralised2019; @goldingGretaGpGaussian2019]. These can also be used to model spatial and teemporal random effects and account for drivers of species vital rates that are not captured in the measured environmental variables. Prespecified mathematical relationships or covariate values can also be included, enabling incorporation of  spatial and temporal predictions from physiological models.

@@

Whilst the model presented in Example 1 makes predictions of spatial variation in vital rates, and provides estimates of uncertainty those predictions, there is no spatially-varying information on vital rates with which to test those assumptions for this species. The time and cost required to robustly estimate how a species' vital rates vary with environmental covariates means that such estimates have been produced for relatively few species - one of the motivations for DSDMs.

@@

Example 1 indicates that identifiability issues do not affect predictions of the probability of presence - even if the data and priors do not provide sufficient information to disentangle vital rates.
This example also suggests that a minimal amount of information is required to make the vital rates identifiable.
These issues of identifiability warrant more detailed investigation, and healthy scepticism on the part of users.
However since identifiability will depend on the nature of the data, model, and prior information, a detailed assessment of model identifiability is beyond the scope of this manuscript.

@@

Bayesian inference on the DSDM for the Carolina Chickadee by MCMC took `r round(bbs_time)` seconds on a laptop computer. For this particular model, the computation time can be sped up by using an analytic solution to the transition matrix eigenvalues: $\lambda_i = F_i S_{J_i} + S_{A_i}$.
Fitting the analytic version of the model with the same Bayesian inference routine took `r round(bbs_analytic_time)` seconds on the same machine.
Such convenient and computationally efficient analytic solutions will not be available for all DSDM models however - particularly where matrices are large or encode more complex demography, such as trait-dependent or temporally stochastic vital rates, density dependence, or dispersal. Moreover, deriving analytic solutions for quantities relating to matrix population models is likely to be outside the skill-set of most ecologists, and in any case would force users to focus on model implementation details rather than the ecology of their study species. Consequently, the ability to efficiently analyse large numbers of population transition matrices during each iteration of the inference algorithm is critical to ensuring DSDMs are a practically useful tool for ecologists.

The greta extension package greta.dynamics provides methods for efficiently analysing multiple population transition matrices to determine their intrinsic growth rates and stable stage distributions. The package also enables efficient analysis of dynamic transition matrices, where vital rates and therefor the matrix at each iteration depend on the previous population size or temporally-varying covariates - enabling modelling of density-dependent or temporally stochastic demographics in DSDMs. Whilst dispersal can already be implemented by convolving population transition matrices with a dispersal matrix, future versions of the R package will facilitate more computationally efficient algorithms for accounting for dispersal - leveraging the inherent sparsity in dispersal matrices.

>>>>

3. Example applications mapping the Carolina Chickadee in North America, and a Protea in southern Africa indicate that DSDMs have similar predictive performance to SDMs, and produce plausible spatial estimates of vital rates. Even when DSDMs are provided little information on the ecology of the target species, predictions of distributions are well-identified, and increased uncertainty about population dynamics is reflected in more uncertain estimates of vital rates.

@@

Whilst users typically have some knowledge of the processes driving these aspects of species distributions &mdash; such as the species' life history, approximate survival and fecundity rates or dispersal ability &mdash; incorporating this knowledge into correlative statistical models is problematic.
On the other hand, ecologists rarely have enough information to accurately predict species distributions from their knowledge of the species life history and demographics alone, since quantifying how vital rates vary over space typically requires large-scale long-term field studies [@merowUsingIntegralProjection2014] or detailed physiological modelling [@kearneyModellingSpeciesDistributions2008].

@@

A number of models have been made to improve the realism in species distribution models, whilst still enabling them to be fitted to the distribution data that are commonly available.
For example, dynamic range models (DRMs) extend correlative SDMs by explicitly modelling dispersal processes and spatial variation in population growth rates [@pagelForecastingSpeciesRanges2012].
DRMs model population growth rates using a correlative model and environmental covariates, just as a correlative SDM models the abundance or probability of presence of a species. 
DRMs can also be fitted to distribution data, though they are intended for use with time series of abundance or presence-only data, since repeat observations are needed to accurately quantify population growth rates and dispersal.

Because they do not model vital rates directly DRMs, like correlative SDMs, are unable to predict the impact on distributions of a change &mdash; for example a conservation intervention like predator control &mdash; that affects a single vital rate.

@@

By contrast, matrix population models and integral projection models (IPMs) are demographic models of single populations that construct and analyse a transition matrix representing key demographic processes [@merowAdvancingPopulationEcology2014].
Spatial IPMs extend this idea to map distributions by modelling demographic rates as a function of covariates, then analysing transition matrices under different conditions to map emergent demographic quantities (such as the intrinsic growth rate) to characterize the distribution of a species [@merowUsingIntegralProjection2014].

@@

where $\mathbf{X}$ is a design matrix of environmental covariates and an intercept dummy variable for the fecundity and survival submodels and $\beta_F$ and $\beta_S$ are vectors of regression coefficients for the two submodels.
Rather than separately modelling spatial variation in adult and juvenile fecundity, I assume that the environmental predictors of survival are the same in both stages, but that juvenile survival differs from adult survival by a constant amount on the logit scale, this maintains the (0,1) constraint on survival probabilities.

@@

where $eigval()$ denotes the dominant eigenvalue of a symmetric matrix, yielding the intrinsic growth rate $\lambda_i$ of the population at location $i$. The theoretical carrying capacity at each site $K_i$ is computed from the carrying capacity relationship of the Beverton-Holt model. $K_i$ has negative values $\lambda_i < 1$, so it is transformed to an always-positive relative abundance parameter $\Lambda_i$ via the softplus function. The softplus transformation is a smooth version of the Heaviside function (which sets negative values to zero), so $\Lambda_i$ closely approximates $K_i$ for most $lambda_i > 1$, and rapidly approaches zero below $lambda_i = 1$. This formulation is statistically convenient, but as discussed below, DSDMs can incorporate more complex and realistic forms of density-dependence if required. 
The parameter $\alpha$ controls the relationship between the intrinsic population growth rate and relative abundance. It is a compound of two parameters that are only identified together: the Beverton-Holt density-dependence parameter relating growth rates to carrying capacities; and the detection rate &mdash; the fraction of individuals in the total population that are detected during a presence-absence survey at site $i$.  
From $\Lambda_i$, we obtain the probability $p_i$ of detecting (one or more individuals of) the species $y_i$, under a poisson sampling distribution [@pagelQuantifyingRangewideVariation2014].

@@

The spatially-averaged fecundity and adult survival estimates from @ryuDevelopingPopulationModels2016 were used to define priors on regression coefficients for the vital rate submodels. These estimated vital rates correspond to populations at the $J$ survey stations used in the recapture analysis (station names provided in the supplemental data to @ryuDevelopingPopulationModels2016) &mdash; locations where the species occurs. I defined a prior on $\beta_F$ and $\beta_S$ such that values were *a priori* more likely if they resulted in estimates of $F_j$ and $S_{A_j}$, for each survey station location $j$, that were more consistent with the estimates in @ryuDevelopingPopulationModels2016.
First, the means and variances of these estimates were used to calculate equivalent parameters of univariate normal distributions on the link-scale of the regression submodels, resulting in the implied priors: $ln(F_j) \sim N(-0.63, 0.18)$, $logit(S_{A_j}) \sim N(1.12, 1.28)$, and $logit(S_{J_j}) \sim N(-0.05, 0.85)$.
These normally-distributed implied priors correspond exactly to multivariate normal priors on the vectors of regression coefficients, since the matrix-multiplication that maps between regression coefficients and link-scale response is an affine transformation, under which the multivariate normal distribution is closed. The parameters of the priors over $\beta_F$ are therefore given by: $\mu_F = \mathbf{X}_*^{-1}(\text{-}0.63\, \mathbf{I})$ and $\mathbf{\Sigma}_F = \mathbf{X}_*^{-1} (0.18\,\mathbf{I}) \mathbf{X}_*^{-T}$, where $\mathbf{X}_*$ denotes the design matrix over the $J$ survey stations, $\mathbf{X}_*^{-1}$ its Moore-Penrose pseudoinverse, $\mathbf{X}_*^{-T}$ the transpose of that pseudoinverse, and $\mathbf{I}$ an identity matrix of dimension $J$. The prior parameters $\mu_S$ and $\mathbf{\Sigma}_S$ were defined in the same way using the implied prior parameters over $logit(S_{A_j})$: 1.12, and 1.28.
Note that these priors do not provide any information on the direction of the relationships between vital rates and environmental covariates (the intercept elements of $\mu_F$ and $\mu_S$ are informed, but the prior mean for the other terms is $\approx 0$), but by providing information on plausible magnitudes of, and correlations between these parameters the prior favours regression equations that pass close to the vital rate estimates at $F_j$ and $S_j$.
The prior for $\gamma$, the logit-difference between adult and juvenile survival, is given by the difference between the implied normal priors on logit-survival of adults and juveniles, which is also normal.
In the absence of any meaningful prior information for the competition effects or detection rates, I assigned $\alpha$ an improper uniform prior on all positive values.

I fitted the model to a random subset of `r bbs_n_train` presence-absence records from the North American Breeding Bird Survey in 2015 [@sauerNorthAmericanBreeding2011]. I used five environmental covariates in $\mathbf{X}$: percentage cover of deciduous forest, coniferous forest, and mixed forest (to represent breeding habitat for this species) mean temperature (to capture thermal limits on survival of the species), and annual precipitation (as a proxy for insect abundance and plant growth, food sources for this species). The forest cover layers were obtained from the maxlike R package [@royleLikelihoodAnalysisSpecies2012], and the climatic covariates were bioclim layers 1 and 12 from the WorldClim 2.0 dataset [@fickWorldClimNew1km2017]. In general, ecological knowledge could be used to select separate covariates to use for each of the submodels. However, in the absence of a strong belief that any of these covariates would affect only one vital rate, all six covariates were used in both submodels. These covariate values were centred and scaled (before calculation of the priors on $\beta_F$ and $\beta_S$) to enable comparison of regression coefficients as effect sizes.

@@

Note that this model could undoubtedly be improved by including more information of the species ecology and vital rates &mdash; for example constraining spatial estimates of fecundity and survival to within a range of plausible values &mdash; via prior distributions or changes to model structure. As with any model, the reliability of these vital rate and distribution predictions will depend on the assumptions made and information provided by the modeller. Unlike in purely correlative models however, these assumptions are made clear by explicitly considering spatially-varying vital rates.

@@

Whilst statistical inference against distribution data could be performed on almost any demographic model, I restrict the models discussed here to those that are amenable to efficient MCMC methods. Specifically: models with likelihoods that are tractable (the exact value of likelihood can be computed from the parameters and data), and can be differentiated with respect to parameter values. This enables the use of modern gradient-based MCMC samplers [@monnahanFasterEstimationBayesian2017], though it necessarily restricts the scope of these models to those with only continuous parameters. The model variations I propose below encompass a wide range of features whilst remaining within these constraints, and can also make use of efficient vectorised and linear algebra routines, meaning that should scale well to the large datasets commonly used for fitting SDMs.

@@

Density dependence can be represented in DSDMs by modelling demographic rates as a function of the current population state [@ellnerIntegralProjectionModels] (the abundances of each life stage) as well as covariates.
For example, the current density of individuals could determine fecundity for the next iteration of the dynamics. The matrix iterations will then converge to a stable state; the population carrying capacity or the expected abundance. this can be achieved by iterating populations a sufficient number of times $K$ for the populations to converge on stable population sizes.

@@

for iteration $k$ in $K$.
 note that this can be combined with Eq 5, or can be iterated for a sufficient number of timesteps ($T$) to get a temporally-static density-dependent model. In practice, $T$ can be determined adaptively, iterating a sufficient number of times for $N_i$ to converge, and then terminating.

@@

For example, these linear models could be applied to very large numbers of covariates and transformations thereof in combination with regularising priors to perform automatic variable selection, similarly to the popular Maxent SDM software[@phillipsModelingSpeciesDistributions2008]. Or to specify mixed-effects models to account for clustered sampling designs, or use random slopes to share environmental reponse information between species in a biotic interaction DSDM. Instead of linear models, non-linear functions could be introduced with generalisd additive models or Gaussian process models, for which there are existing greta extension packages: greta.gam and greta.gp [@goldingGretaGamGeneralised2019; @goldingGretaGpGaussian2019]. These can also be used to model spatial and temporal random effects and account for drivers of species vital rates that are not captured in the measured environmental variables. Prespecified mathematical relationships or covariate values can also be included, enabling incorporation of  spatial and temporal predictions from physiological models.

@@

Whilst the model presented in Example 1 makes predictions of spatial variation in vital rates, and provides estimates of uncertainty in those predictions, there is no spatially-varying information on vital rates with which to test those predictions for this species. The time and cost required to robustly estimate how a species' vital rates vary with environmental covariates means that such estimates have been produced for relatively few species; one of the motivations for DSDMs.

@@

Example 1 indicates that identifiability issues do not affect predictions of the probability of presence, even if the data and priors do not provide sufficient information to disentangle vital rates.
This example also suggests that a minimal amount of information is required to make the vital rates identifiable.
These issues of identifiability warrant more detailed investigation, and healthy scepticism on the part of users.
However since identifiability will depend on the nature of the data, model, and prior information, a detailed assessment of model identifiability is beyond the scope of this manuscript.

@@

Bayesian inference on the DSDM for the Carolina Chickadee by MCMC took `r round(bbs_time)` seconds on a laptop computer. For this particular model, the computation time can be sped up by using an analytic solution to the transition matrix eigenvalues: $\lambda_i = F_i S_{J_i} + S_{A_i}$.
Fitting the analytic version of the model with the same Bayesian inference routine took `r round(bbs_analytic_time)` seconds on the same machine.
Such convenient and computationally efficient analytic solutions will not be available for all DSDM models, however,particularly where matrices are large or encode more complex demography, such as trait-dependent or temporally stochastic vital rates, density dependence, or dispersal. Moreover, deriving analytic solutions for quantities relating to matrix population models is likely to be outside the skill-set of most ecologists, and in any case would force users to focus on model implementation details rather than the ecology of their study species. Consequently, the ability to efficiently analyse large numbers of population transition matrices during each iteration of the inference algorithm is critical to ensuring DSDMs are a practically useful tool for ecologists.

The greta extension package greta.dynamics provides methods for efficiently analysing multiple population transition matrices to determine their intrinsic growth rates and stable stage distributions. The package also enables efficient analysis of dynamic transition matrices, where vital rates and therefore the matrix at each iteration depend on the previous population size or temporally-varying covariates,enabling modelling of density-dependent or temporally stochastic demographics in DSDMs. Whilst dispersal can already be implemented by convolving population transition matrices with a dispersal matrix, future versions of the R package will facilitate more computationally efficient algorithms for accounting for dispersal, leveraging the inherent sparsity in dispersal matrices.
